Viscous Transport in Eroding Porous Media

### 2020-01-27 docs/transport/Response_to_Reviewer_1.tex

\documentclass[11pt]{article}

\usepackage{todonotes}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb,stmaryrd}
\usepackage{color}
\newcommand{\comment}[1]{{\color{blue} #1}}

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography, and minor grammatical errors
were corrected throughout. An itemized list of the changes addressing
the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

@@

    \begin{quotation}
      \noindent
      Section 5 describes the numerical methods, with special attention
      paid to the quadrature method for computing the shear stress and
      the vorticity.
    \end{quotation}

    \begin{quotation}
      \noindent
      As a continuation of our previous work (Quaife \& Moore 2018), we
      have simulated dense suspensions and characterized transport in
      viscous eroding porous media. This is accomplished by using
      high-order time stepping methods and a combination of trapezoid
      and Barycentric quadrature methods to solve a BIE formulation of
      the Stokes equations. By using these numerical methods, we are
      able to perform stable simulations of erosion with $N = O(100)$
      discretization points, while the trapezoid rule would require
      $O(10^5)$ discretization points.
    \end{quotation}
\end{itemize}

\noindent
\comment{The transient increase of tortuosity is interesting, but the
explanation in Sec 6.4 is not satisfying. The authors proposed that the
increase of tortuosity can be explained by the sudden increase of
$x$-velocity when a grain vanishes. But in formula (4.4) or (4.5), the
$x$-velocity appears both in the numerator and the denominator, which
seems to me that the increase would just cancel. Then, as the authors
have pointed out, wouldn't the straighter streamlines resulted from the
vanishing of grains have decreased the tortuosity?}
\begin{itemize}
  \item This was an oversight on our behalf. The jump is actually caused
    by a grain completely vanishing, but not for the reason explained in
    the original submission.

  \item At the large porosity, there is a long chain of small bodies
    that, relative to the rest of the geometry, creates a region of
    large local tortuosity. When the left-most grain completely erodes,
    this slightly changes the local tortuosity, but more importantly, it
    increases the inlet velocity relative to the remainder of the
    channel. Since the tortuosity is the integral of the local
    tortuosity weighted by the initial velocity, this increase in
    velocity creates a slight increase in the tortuosity.

  \item We have updated the explanation to read:
    \begin{quotation}
      \noindent
      For this geometry, the tortuosity decreases monotonically at
      almost all porosities. However, the tortuosity slightly increases
      near the end of the simulation. This increase is caused when a
      grain located at the start of a region with large local tortuosity
      completely erodes.  In this region, once the grain vanishes, the
      large local tortuosity changes slightly, but the inlet velocity,
      $u_1(y)$, relative to the inlet velocity in the remainder of the
      channel, increases sharply. The result is a sudden increase in the
      average local tortuosity when weighted by the initial velocity,
      and this increases the tortuosity of the geometry
    \end{quotation}
\end{itemize}

@@

\noindent
\comment{The porosity-tortuosity models (4.6c \& 4.6d) are not
explicitly compared in the results section. How much worse are they
comparing to the other models?}
\begin{itemize}
  \item We have adjusted each of the appropriate sections.

  \item In Section 6.2, we have added the following sentence:
    \begin{quotation}
      \noindent
      A slightly larger root-mean-square error of $6.70 \times 10^{-3}$
      is possible with the logarithmic porosity-tortuosity model (4.6b),
      but the other two models have a root- mean-square error that is an
      order of magnitude larger.
    \end{quotation}

  \item In Section 6.3, we have added the following sentence:
    \begin{quotation}
      \noindent
      However, the root-mean-square error of the other three models is
      bounded by $2.38 \times 10^{-2}$.
    \end{quotation}

  \item In Section 6.4, we have added the following sentence:
    \begin{quotation}
      \noindent
      We compute the lines of best fit using the porosity-tortuosity
      models (4.6).  Each of the models result in a root-mean-squared
      error ranging from $4.60 \times 10^{-3}$ for model (4.6d) to $7.20
      \times 10^{-3}$ for model (4.6c). The black dashed line in figure
      15(a) is the line of best fit $\widehat{T}(\phi) =
      \phi^{-0.2459}$, with a root-mean-square error of $5.50 \times
      10^{-3}$.  We note a slightly better root-mean-square error of
      $4.60 \times 10^{-3}$ is possible with the model
      $\widehat{T}(\phi) = \left(1 + 0.1499(1-\phi)\right)^2$.
    \end{quotation}

@@

\noindent
\comment{``Porosity" is undefined throughout the paper.}
\begin{itemize}
  \item We have now formally defined the porosity by rewriting the
    description of coarse-grained models as:
  \begin{quotation}
  \noindent
    Examples of coarse-grained models for porous media flow include
    tortuosities (Matyka {\em et al.}~2008; Duda {\em et al.}~2011;
    Koponen {\em et al.}~1996), geometry connectivity (Knudby \& Carrera
    2005), anomalous dispersion (Dentz {\em et al.}~2004), and
    permeability-porosity relationships (Dardis \& McCloskey 1998;
    Carman 1937), where the porosity is the fraction of the geometry
    that is fluidized.
  \end{quotation}

\end{itemize}

\noindent
\comment{``Area fraction" is plotted in (and only in) Figure 7, while
the rest of the paper used porosity. Maybe plotting porosity here would
give a more consistent and smoother reading experience?}
\begin{itemize}
  \item We examined the area fraction and flow rate in figure 7 so that
    it could be compared to results from our previous work. The goal is
    to demonstrate that we now are able to resolve smaller area
    fractions and correspondingly lower flow rates.

  \item We have updated the text to read:
    \begin{quotation}
      \noindent
      In our previous work (Quaife \& Moore 2018), we investigated the
      effects of erosion on the area fraction and flow rate of two
      different geometries. Here, we do the same analysis, but with the
      denser initial packing of the grains.
    \end{quotation}

### 2020-01-27 docs/transport/erosion_transport.tex
finished up reviews

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared with models available in
  the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.
\end{abstract}

@@

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains. In previous
work~\citep{qua-moo2018}, the BIE is discretized with the trapezoid
rule, and this is acceptable if the grains are sufficiently
separated~\citep{tre-wei2014}. However, when grains are close, the
result is a nearly-singular integrand and the error of the trapezoid
rule is no longer acceptable. One of the earliest quadrature methods for
nearly-singular integrands was developed by~\citet{bak-she1986}, and in
recent years, many other schemes have followed~\citep{kli-tor2018,
hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We
use a Barycentric quadrature method~\citep{bar2014, bar-wu-vee2015}
since it is a non-intrusive modification of the trapezoid rule, and the
error is guaranteed to be uniformly bounded.  We use this quadrature
method to compute the velocity gradient, which is needed to evaluate the
shear stress and the fluid vorticity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. Finally, we define the pore
sizes to be the minimum distance between all pairs of neighboring
grains, and these distances are represented by the red line segments in
figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}. In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes, represented by
the length of the red line segments, between neighboring grains. Two
grains are neighbors if they share an edge of the Delanuay triangulation
with nodes at the center of each eroding grain. (b) The pore sizes of an
eroded geometry. The black curve is the Weibull distribution with the
same first two moments as the data.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using these quadrature rules, the velocity, shear stress, and
vorticity are calculated at the target points to simulate erosion and
transport. However, applied directly, it requires $O(N^2)$ operations,
where $N$ is the total number of source and target points.  By using a
fast summation method, such as the fast multipole method
(FMM)~\citep{gre-rok1987}, the cost can be reduced to $O(N)$ operations.
However, each application of the Barycentric quadrature rules involve
several $N$-body calculations, rendering the computational cost
prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and an accelerated trapezoid rule.  Note
that the source points of the layer potential is always one of the
eroding bodies or the outer wall, but the target point can either be on
another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Single Body Close to a Wall}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Consider a single eroding body close to a solid wall with the shear flow
$\UU(\xx) = (y+1,0)$ imposed on $\Gamma$. \citet{mit-spa2017} performed
a similar three-dimensional numerical experiment using a second-order
quadrature method.  Their initial body is a sphere with its center
located $1.5$ radii above the solid wall.  We initialize the
two-dimensional eroding body with radius $r = 0.4$, and we conduct
numerical experiments where the initial distance between the grain and
the solid wall is $h$, $h/2$, and $h/10$, where $h = 2\upi r/N_\iin$.
If we used the trapezoid rule and required an error that is comparable
to the Barycentric quadrature rule, the body with an initial distance of
$h/10$ from the solid wall would require $6,400$ discretization points,
and the outer wall would require $50,000$ discretization points.

@@

\begin{figure}
\begin{center}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r1h_shear}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r0d5h_shear}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r0d1h_shear}
\caption{}
\end{subfigure}
\end{center}
\caption{\label{fig:NearWall} A single body eroding in a shearing Stokes
flow.  The color is the logarithm of the shear stress. Therefore,
erosion is fastest in the red regions (upper half) and slowest in the
blue regions (lower half).  The body is initialized at three different
distances from the lower wall: (a) $h$, (b) $h/2$, and (c) $h/10$.}
\end{figure}
\begin{figure}
\begin{center}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r1h_vort}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r0d5h_vort}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r0d1h_vort}
\caption{}
\end{subfigure}
\caption{\label{fig:NearWall_vort} The eroded bodies from
figure~\ref{fig:NearWall} at time $t=0.1$. The color is the logarithm of
the magnitude of the vorticity. The initial distances from the body to
the solid wall are: (a) $h$, (b) $h/2$, and (c) $h/10$.}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We randomly initialize 20 eroding grains with an initial porosity of
$\phi = 37.68\%$ and the Hagen-Poiseuille flow $\UU(\xx) = U
\left(1-y^2,0 \right)$ imposed on $\Gamma$. The flow rate $U$ is chosen
so that the average pressure drop from $x=-2$ to $x=2$ is held fixed at
8. Therefore, $U=1$ once all the grains have vanished.  The vorticity
and grain configuration at four equispaced times are shown in
figure~\ref{fig:Eroding20vort}.  Initially, several of the grains are
closer to the outer wall than the $5h$ threshold required for the
trapezoid rule to achieve machine precision.  In particular, the
distance between bodies 1, 6, 13, and 15 and the outer wall is $1.3h$,
$2.9h$, $2.8h$, and $1.3h$, respectively, where $h$ is the arclength
spacing of the outer wall $\Gamma$.  In addition, the distance between
several pairs of eroding bodies, including 1 \& 6, 3 \& 9, 6 \& 8, and
14 \& 18, is too small to be resolve with the trapezoid rule.  By using
the Barycentric quadrature rule, the interaction between these
nearly-touching bodies is resolved to the desired accuracy, and erosion
can be simulated until all the bodies have vanished.

@@

In our previous work~\citep{qua-moo2018}, we investigated the effects
of erosion on the area fraction and flow rate of two different
geometries.  Here, we do the same analysis, but with the denser packing
of the grains. In figure~\ref{fig:Eroding20flowrate}(a), we plot the
area fraction as a function of normalized time.  The general trend of
the area fraction resembles our previous work~\citep[see][figure
10(a)]{qua-moo2018}, but with a larger initial area fraction.  In
figure~\ref{fig:Eroding20flowrate}(b), we plot the flow rate $U$
required to maintain a constant pressure drop across the channel.
Again, the trend of $U$ resembles that of our previous
work~\citep[see][figure 10(b)]{qua-moo2018}, except that the initial
flow rate is an order of magnitude smaller because of the larger initial
area fraction.  Starting around normalized time $0.2$,
figure~\ref{fig:Eroding20flowrate}(b) is roughly linear which indicates
that the flow rate can be written as an exponential law.  The line of
best fit is $U \approx \exp(9.94(t-t_f)/t_f)$ which is the dashed line
in figure~\ref{fig:Eroding20flowrate}(b).
\begin{figure}
\begin{center}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b30n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b90n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b150n}\\
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b210n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270n_zoom}
\caption{\label{fig:Eroding20tracer} 200 particle trajectories in the
second geometry from figure~\ref{fig:Eroding20vort}. The particles are
initially equispaced at $(-1,y)$, where $y \in (-1,1)$. The first five
snapshots are equispaced in time.  The bottom right frame is a
magnification of the fifth snapshot, but with additional trajectories.}
\end{center}
\end{figure}

In figure~\ref{fig:Eroding20vort}, we observe that erosion creates a
network of channels from the inlet to the outlet where the velocity and
vorticity, and therefore erosion rate, are much larger relative to other
regions.  These channels can be further visualized with particle
trajectories that are trace out the streamlines.  In
figure~\ref{fig:Eroding20tracer}, we freeze the geometry at the second
snapshot from figure~\ref{fig:Eroding20vort} and plot 200 trajectories
that are initially equispaced along $(-1,y)$, where $y \in (-1,1)$.  The
trajectories are shown at five different times, and the final plot is a
zoom in of the lower right quadrant of the fifth time step, but with
additional trajectories.  Since we use a high-order quadrature rule and
time stepping method, we resolve the dynamics very close to the eroded
bodies.  There are three clear regions where the streamlines are most
concentrated, corresponding to the regions of highest velocity.  Two of
these regions are located between the bodies and the solid walls at
$y=\pm 1$, and the third cuts through the porous region with the upper
part of the channel formed by bodies 1, 4, 6, and 8.  Since the flow is
fastest in these regions, the shearing is largest, and this causes the
channels to continue to open fastest as observed in
figure~\ref{fig:Eroding20vort}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with of 20 randomly initialized eroding
bodies, but with a smaller initial porosity of $\phi = 30.67\%$.  In
figure~\ref{fig:ErodingLow20vort}, we plot the eroding geometry and
vorticity at four evenly spaced instances in time.  Initially, the
smallest distance between pairs of bodies is $3.29 \times 10^{-4}$, and
the smallest distance between the bodies and solid wall is $4.50 \times
10^{-3}$.  At these distances, a resolution of approximately $N_\iin =
27,000$ and $N_\out = 18,000$ discretization points is required to
satisfy the $5h$ threshold needed for the trapezoid rule to achieve
machine precision.

>>>>

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared with models available in
  the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a boundary integral equation solver for the fluid equations,
  a second-order stable time stepping method to simulate erosion, and
  high-order numerical methods to stably and accurately resolve
  nearly-touching eroded bodies and particle trajectories near the
  eroding bodies.
\end{abstract}

@@

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains. In previous
work~\citep{qua-moo2018}, the BIE is discretized with the trapezoid
rule, and this is acceptable if the grains are sufficiently
separated~\citep{tre-wei2014}. However, when grains are close, the
result is a nearly-singular integrand and the error of the trapezoid
rule is no longer acceptable. One of the earliest quadrature methods for
nearly-singular integrands was developed by~\citet{bak-she1986}, and in
recent years, many other schemes have followed~\citep{kli-tor2018,
hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We
use a Barycentric quadrature method~\citep{bar2014, bar-wu-vee2015}
since it is a non-intrusive modification of the trapezoid rule, and the
error is guaranteed to be uniformly bounded.  We use this quadrature
method to compute the velocity gradient, which is needed to evaluate the
shear stress and the fluid vorticity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes, represented by
the length of the red line segments, between neighboring grains. Two
grains are neighbors if they share an edge of the Delanuay triangulation
with nodes at the center of each eroding grain. (b) The pore sizes of an
eroded geometry. The black curve is the Weibull distribution with the
same first two moments as the data.}
\end{figure}

While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. Finally, we define the pore
sizes to be the minimum distance between all pairs of neighboring
grains, and these distances are represented by the red line segments in
figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}. In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using these quadrature rules, the velocity, shear stress, and
vorticity are calculated at the target points to simulate erosion and
transport. However, applied directly, it requires $O(N^2)$ operations,
where $N$ is the total number of source and target points.  By using a
fast summation method, such as the fast multipole method
(FMM)~\citep{gre-rok1987}, the cost can be reduced to $O(N)$ operations.
However, each application of the Barycentric quadrature rules involve
several $N$-body calculations, rendering the computational cost
prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and an accelerated trapezoid rule.  Note
that the source points of the layer potential are always one of the
eroding bodies or the outer wall, but the target point can either be on
another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Single Body Close to a Wall}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{center}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r1h_shear}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r0d5h_shear}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.27\textwidth}
\includegraphics[height = \textwidth]{./figs/1b_0d4r0d1h_shear}
\caption{}
\end{subfigure}
\end{center}
\caption{\label{fig:NearWall} A single body eroding in a shearing Stokes
flow.  The color is the logarithm of the shear stress. Therefore,
erosion is fastest in the red regions (upper half) and slowest in the
blue regions (lower half).  The body is initialized at three different
distances from the lower wall: (a) $h$, (b) $h/2$, and (c) $h/10$.}
\end{figure}
\begin{figure}
\begin{center}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r1h_vort}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r0d5h_vort}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.32\textwidth}
\includegraphics[height = 0.53\textwidth]{./figs/1b_0d4r0d1h_vort}
\caption{}
\end{subfigure}
\caption{\label{fig:NearWall_vort} The eroded bodies from
figure~\ref{fig:NearWall} at time $t=0.1$. The color is the logarithm of
the magnitude of the vorticity. The initial distances from the body to
the solid wall are: (a) $h$, (b) $h/2$, and (c) $h/10$.}
\end{center}
\end{figure}

Consider a single eroding body close to a solid wall with the shear flow
$\UU(\xx) = (y+1,0)$ imposed on $\Gamma$. \citet{mit-spa2017} performed
a similar three-dimensional numerical experiment using a second-order
quadrature method.  Their initial body is a sphere with its center
located $1.5$ radii above the solid wall.  We initialize the
two-dimensional eroding body with radius $r = 0.4$, and we conduct
numerical experiments where the initial distance between the grain and
the solid wall is $h$, $h/2$, and $h/10$, where $h = 2\upi r/N_\iin$.
If we used the trapezoid rule and required an error that is comparable
to the Barycentric quadrature rule, the body with an initial distance of
$h/10$ from the solid wall would require $6,400$ discretization points,
and the outer wall would require $50,000$ discretization points.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We randomly initialize 20 eroding grains with an initial porosity of
$\phi = 37.68\%$ and the Hagen-Poiseuille flow $\UU(\xx) = U
\left(1-y^2,0 \right)$ imposed on $\Gamma$. The flow rate $U$ is chosen
so that the average pressure drop from $x=-2$ to $x=2$ is held fixed at
8. Therefore, $U=1$ once all the grains have vanished.  The vorticity
and grain configuration at four equispaced times are shown in
figure~\ref{fig:Eroding20vort}.  Initially, several of the grains are
closer to the outer wall than the $5h$ threshold required for the
trapezoid rule to achieve machine precision.  In particular, the
distance between bodies 1, 6, 13, and 15 and the outer wall is $1.3h$,
$2.9h$, $2.8h$, and $1.3h$, respectively, where $h$ is the arclength
spacing of the outer wall $\Gamma$.  In addition, the distance between
several pairs of eroding bodies, including 1 \& 6, 3 \& 9, 6 \& 8, and
14 \& 18, is too small to resolve with the trapezoid rule.  By using the
Barycentric quadrature rule, the interaction between these
nearly-touching bodies is resolved to the desired accuracy, and erosion
can be simulated until all the bodies have vanished.

@@

In our previous work~\citep{qua-moo2018}, we investigated the effects of
erosion on the area fraction and flow rate of two different geometries.
Here, we do the same analysis, but with the denser initial packing of
the grains. In figure~\ref{fig:Eroding20flowrate}(a), we plot the area
fraction as a function of normalized time.  The general trend of the
area fraction resembles our previous work~\citep[see][figure
10(a)]{qua-moo2018}, but with a larger initial area fraction.  In
figure~\ref{fig:Eroding20flowrate}(b), we plot the flow rate $U$
required to maintain a constant pressure drop across the channel.
Again, the trend of $U$ resembles that of our previous
work~\citep[see][figure 10(b)]{qua-moo2018}, except that the initial
flow rate is an order of magnitude smaller because of the larger initial
area fraction.  Starting around normalized time $0.2$,
figure~\ref{fig:Eroding20flowrate}(b) is roughly linear which indicates
that the flow rate can be written as an exponential law.  The line of
best fit is $U \approx \exp(9.94(t-t_f)/t_f)$ which is the dashed line
in figure~\ref{fig:Eroding20flowrate}(b).
\begin{figure}
\begin{center}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b30n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b90n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b150n}\\
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b210n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270n}
\includegraphics[width = 0.32 \textwidth]{./figs/tracer_20b270n_zoom}
\caption{\label{fig:Eroding20tracer} 200 particle trajectories in the
second geometry from figure~\ref{fig:Eroding20vort}. The particles are
initially equispaced at $(-1,y)$, where $y \in (-1,1)$. The first five
snapshots are equispaced in time.  The bottom right frame is a
magnification of the fifth snapshot, but with additional trajectories.}
\end{center}
\end{figure}

In figure~\ref{fig:Eroding20vort}, we observe that erosion creates a
network of channels from the inlet to the outlet where the velocity and
vorticity, and therefore erosion rate, are much larger relative to other
regions.  These channels can be further visualized with particle
trajectories that trace out the streamlines.  In
figure~\ref{fig:Eroding20tracer}, we freeze the geometry at the second
snapshot from figure~\ref{fig:Eroding20vort} and plot 200 trajectories
that are initially equispaced along $(-1,y)$, where $y \in (-1,1)$.  The
trajectories are shown at five different times, and the final plot is a
zoom in of the lower right quadrant of the fifth time step, but with
additional trajectories.  Since we use a high-order quadrature rule and
time stepping method, we resolve the dynamics very close to the eroded
bodies.  There are three clear regions where the streamlines are most
concentrated, corresponding to the regions of highest velocity.  Two of
these regions are located between the bodies and the solid walls at
$y=\pm 1$, and the third cuts through the porous region with the upper
part of the channel formed by bodies 1, 4, 6, and 8.  Since the flow is
fastest in these regions, the shearing is largest, and this causes the
channels to continue to open fastest as observed in
figure~\ref{fig:Eroding20vort}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with 20 randomly initialized eroding
bodies, but with a smaller initial porosity of $\phi = 30.67\%$.  In
figure~\ref{fig:ErodingLow20vort}, we plot the eroding geometry and
vorticity at four evenly spaced instances in time.  Initially, the
smallest distance between pairs of bodies is $3.29 \times 10^{-4}$, and
the smallest distance between the bodies and solid wall is $4.50 \times
10^{-3}$.  At these distances, a resolution of approximately $N_\iin =
27,000$ and $N_\out = 18,000$ discretization points is required to
satisfy the $5h$ threshold needed for the trapezoid rule to achieve
machine precision.
